---
title: Campus Health Centre Demand Analytics
author: Sneha Bhowmick
date: August 10, 2026
output: html_document
---

# 1. Project Theme

## Campus Health Centre Demand Analytics

The project aims to use healthcare appointment data to understand and predict patient attendance patterns and thereby support estimation of patient load at a campus health centre.

# 2. Selected Type of Analytics

## Predictive Analytics

Predictive analytics has been selected as the initial type of analytics to investigate.

The objective is to use historical appointment and patient information to predict whether a scheduled patient is likely to attend the appointment. The resulting predictions can help estimate the expected patient load and support better planning of health-centre resources.

# 3. Research Question

The specific question for the predictive analysis is:

> Can historical appointment and patient information be used to predict whether a scheduled patient will attend the appointment, thereby helping a campus health centre estimate its expected patient load?

# 4. Dataset Required

A publicly available healthcare appointment dataset containing a large number of historical appointment records is required to investigate the research question.

The proposed dataset is the **Medical Appointment No Shows** dataset. It contains approximately 110,000 appointment records and includes information such as:

- Patient demographic information
- Scheduled date
- Appointment date
- Health-related indicators
- SMS reminder information
- Appointment attendance outcome

The appointment attendance outcome will be used as the target variable for the initial predictive analysis.

# 5. Data Access Plan

The dataset will be accessed from a publicly available online data repository and downloaded in CSV format.

The dataset will then be imported into R using R Markdown. The initial data preparation will include:

1. Examining the number of observations and variables.
2. Understanding the meaning and data type of each variable.
3. Checking for missing values.
4. Checking for duplicate records.
5. Converting date variables into appropriate date formats.
6. Creating useful derived variables from the existing data where appropriate.
7. Selecting relevant variables for further analysis.

The cleaned dataset will subsequently be used for exploratory analysis and predictive modelling.

# 6. Current Status

The project has been initialized.

The analytics type, research question, required dataset and data access plan have been identified.

The next stage will be to obtain and inspect the selected dataset.
# 7. Loading the Dataset

```{r load-data}
data <- read.csv("KaggleV2-May-2016.csv",
                 stringsAsFactors = FALSE)

head(data)

# 8. Initial Data Inspection

The dataset is first inspected to understand its structure, variables, dimensions, and basic statistical characteristics.

```{r data-inspection}
str(data)
summary(data)
dim(data)
names(data)

# 9. Data Quality Check

The dataset is checked for missing values and duplicate records before further analysis.

```{r data-quality}
colSums(is.na(data))
sum(duplicated(data))

# 10. Checking for Invalid Values

The dataset is examined for values that may be outside the expected range of the variables.

```{r invalid-values}
sum(data$Age < 0)
sum(data$Age > 100)

table(data$Handcap)
table(data$Scholarship)
table(data$Hipertension)
table(data$Diabetes)
table(data$Alcoholism)
table(data$SMS_received)
table(data$No.show)

# 11. Data Cleaning and Variable Preparation

The data is prepared for further analysis by handling invalid age values and converting the appointment date variables into appropriate date formats.

```{r data-cleaning}

# Create a cleaned copy of the dataset
data_clean <- data

# Replace the invalid age value of -1 with NA
data_clean$Age[data_clean$Age < 0] <- NA

# Convert date variables to Date format
data_clean$ScheduledDay <- as.Date(data_clean$ScheduledDay)
data_clean$AppointmentDay <- as.Date(data_clean$AppointmentDay)

# Create the waiting time variable
data_clean$WaitingDays <- as.numeric(
  data_clean$AppointmentDay - data_clean$ScheduledDay
)

# Check the cleaned data
summary(data_clean$Age)
summary(data_clean$WaitingDays)

# 12. Checking Waiting-Time Anomalies

Some records may contain an appointment date that occurs before the scheduled date. These records are identified and inspected before deciding how to handle them.

```{r waiting-time-check}

# Count records with negative waiting time
sum(data_clean$WaitingDays < 0)

# Display the records with negative waiting time
data_clean[data_clean$WaitingDays < 0,
           c("ScheduledDay", "AppointmentDay", "WaitingDays")]

# 13. Data Cleaning

The dataset is cleaned by handling invalid age values and inconsistent waiting periods. A separate cleaned dataset is created so that the original dataset remains unchanged.

```{r data-cleaning-final}

# Create a copy of the original dataset
data_clean <- data

# Replace invalid age values with NA
data_clean$Age[data_clean$Age < 0 | data_clean$Age > 100] <- NA

# Convert date variables to Date format
data_clean$ScheduledDay <- as.Date(data_clean$ScheduledDay)
data_clean$AppointmentDay <- as.Date(data_clean$AppointmentDay)

# Calculate waiting time in days
data_clean$WaitingDays <- as.numeric(
  data_clean$AppointmentDay - data_clean$ScheduledDay
)

# Remove records with negative waiting time
data_clean <- data_clean[data_clean$WaitingDays >= 0, ]

# Check the cleaned variables
summary(data_clean$Age)
summary(data_clean$WaitingDays)

```

# 14. Creating Variables for Demand Analysis

Additional variables are created from the appointment date and waiting period to identify patterns in appointment demand.

```{r create-demand-variables}

# Day of the week of the appointment
data_clean$AppointmentDayOfWeek <- weekdays(data_clean$AppointmentDay)

# Month of the appointment
data_clean$AppointmentMonth <- format(
  data_clean$AppointmentDay, "%B"
)

# Week number of the appointment
data_clean$AppointmentWeek <- format(
  data_clean$AppointmentDay, "%U"
)

# Display the newly created variables
head(data_clean[, c(
  "AppointmentDay",
  "AppointmentDayOfWeek",
  "AppointmentMonth",
  "AppointmentWeek",
  "WaitingDays"
)])

```

# 15. Appointment Demand by Day of the Week

The number of appointments is examined across different days of the week to identify patterns in healthcare demand.

```{r demand-by-day}

# Count appointments by day of the week
day_demand <- table(data_clean$AppointmentDayOfWeek)

# Display the appointment counts
day_demand


```
# 16. Visualization of Appointment Demand by Day

The distribution of appointments across the days of the week is visualized to identify the days with higher healthcare demand and potential variations in patient demand throughout the week.

```{r demand-day-plot}

# Set the days in the correct weekly order
day_demand_df <- data.frame(
  Day = c("Monday", "Tuesday", "Wednesday",
          "Thursday", "Friday", "Saturday"),
  Appointments = c(
    day_demand["Monday"],
    day_demand["Tuesday"],
    day_demand["Wednesday"],
    day_demand["Thursday"],
    day_demand["Friday"],
    day_demand["Saturday"]
  )
)

# Create the bar chart
barplot(
  day_demand_df$Appointments,
  names.arg = day_demand_df$Day,
  main = "Appointment Demand by Day of the Week",
  xlab = "Day of the Week",
  ylab = "Number of Appointments",
  las = 2
)
```

**Interpretation:** The visualization shows that appointment demand varies across the days of the week. Wednesday records the highest number of appointments, closely followed by Tuesday. Monday also shows relatively high demand, while Friday and Thursday have fewer appointments. Saturday has an exceptionally low number of appointments. This suggests that patient demand is concentrated mainly during weekdays, particularly in the middle of the week. These findings may help the campus health centre plan staff availability and allocate resources more effectively during high-demand days.