---
title: Campus Health Centre Demand Analytics
author: Sneha Bhowmick
date: August 10, 2026
output: html_document
---

# 1. Project Theme

## Campus Health Centre Demand Analytics

The project aims to use healthcare appointment data to understand and predict patient attendance patterns and thereby support estimation of patient load at a campus health centre.

# 2. Selected Type of Analytics

## Predictive Analytics

Predictive analytics has been selected as the initial type of analytics to investigate.

The objective is to use historical appointment and patient information to predict whether a scheduled patient is likely to attend the appointment. The resulting predictions can help estimate the expected patient load and support better planning of health-centre resources.

# 3. Research Question

The specific question for the predictive analysis is:

> Can historical appointment and patient information be used to predict whether a scheduled patient will attend the appointment, thereby helping a campus health centre estimate its expected patient load?

# 4. Dataset Required

A publicly available healthcare appointment dataset containing a large number of historical appointment records is required to investigate the research question.

The proposed dataset is the **Medical Appointment No Shows** dataset. It contains approximately 110,000 appointment records and includes information such as:

- Patient demographic information
- Scheduled date
- Appointment date
- Health-related indicators
- SMS reminder information
- Appointment attendance outcome

The appointment attendance outcome will be used as the target variable for the initial predictive analysis.

# 5. Data Access Plan

The dataset will be accessed from a publicly available online data repository and downloaded in CSV format.

The dataset will then be imported into R using R Markdown. The initial data preparation will include:

1. Examining the number of observations and variables.
2. Understanding the meaning and data type of each variable.
3. Checking for missing values.
4. Checking for duplicate records.
5. Converting date variables into appropriate date formats.
6. Creating useful derived variables from the existing data where appropriate.
7. Selecting relevant variables for further analysis.

The cleaned dataset will subsequently be used for exploratory analysis and predictive modelling.

# 6. Current Status

The project has been initialized.

The analytics type, research question, required dataset and data access plan have been identified.

The next stage will be to obtain and inspect the selected dataset.
# 7. Loading the Dataset

```{r load-data}
data <- read.csv("KaggleV2-May-2016.csv",
                 stringsAsFactors = FALSE)

head(data)
```

# 8. Initial Data Inspection

The dataset is first inspected to understand its structure, variables, dimensions, and basic statistical characteristics.

```{r data-inspection}
str(data)
summary(data)
dim(data)
names(data)
```

# 9. Data Quality Check

The dataset is checked for missing values and duplicate records before further analysis.

```{r data-quality}
colSums(is.na(data))
sum(duplicated(data))
```

# 10. Checking for Invalid Values

The dataset is examined for values that may be outside the expected range of the variables.

```{r invalid-values}
sum(data$Age < 0)
sum(data$Age > 100)

table(data$Handcap)
table(data$Scholarship)
table(data$Hipertension)
table(data$Diabetes)
table(data$Alcoholism)
table(data$SMS_received)
table(data$No.show)
```
# 11. Data Cleaning and Variable Preparation

The data is prepared for further analysis by handling invalid age values and converting the appointment date variables into appropriate date formats.

```{r data-cleaning}

# Create a cleaned copy of the dataset
data_clean <- data

# Replace the invalid age value of -1 with NA
data_clean$Age[data_clean$Age < 0] <- NA

# Convert date variables to Date format
data_clean$ScheduledDay <- as.Date(data_clean$ScheduledDay)
data_clean$AppointmentDay <- as.Date(data_clean$AppointmentDay)

# Create the waiting time variable
data_clean$WaitingDays <- as.numeric(
  data_clean$AppointmentDay - data_clean$ScheduledDay
)

# Check the cleaned data
summary(data_clean$Age)
summary(data_clean$WaitingDays)
```

# 12. Checking Waiting-Time Anomalies

Some records may contain an appointment date that occurs before the scheduled date. These records are identified and inspected before deciding how to handle them.

```{r waiting-time-check}

# Count records with negative waiting time
sum(data_clean$WaitingDays < 0)

# Display the records with negative waiting time
data_clean[data_clean$WaitingDays < 0,
           c("ScheduledDay", "AppointmentDay", "WaitingDays")]
```

# 13. Data Cleaning

The dataset is cleaned by handling invalid age values and inconsistent waiting periods. A separate cleaned dataset is created so that the original dataset remains unchanged.

```{r data-cleaning-final}

# Create a copy of the original dataset
data_clean <- data
```

# Replace invalid age values with NA
```{r cleaning}
data_clean$Age[data_clean$Age < 0 | data_clean$Age > 100] <- NA

# Convert date variables to Date format
data_clean$ScheduledDay <- as.Date(data_clean$ScheduledDay)
data_clean$AppointmentDay <- as.Date(data_clean$AppointmentDay)

# Calculate waiting time in days
data_clean$WaitingDays <- as.numeric(
  data_clean$AppointmentDay - data_clean$ScheduledDay
)

# Remove records with negative waiting time
data_clean <- data_clean[data_clean$WaitingDays >= 0, ]

# Check the cleaned variables
summary(data_clean$Age)
summary(data_clean$WaitingDays)

```

# 14. Creating Variables for Demand Analysis

Additional variables are created from the appointment date and waiting period to identify patterns in appointment demand.

```{r create-demand-variables}

# Day of the week of the appointment
data_clean$AppointmentDayOfWeek <- weekdays(data_clean$AppointmentDay)

# Month of the appointment
data_clean$AppointmentMonth <- format(
  data_clean$AppointmentDay, "%B"
)

# Week number of the appointment
data_clean$AppointmentWeek <- format(
  data_clean$AppointmentDay, "%U"
)

# Display the newly created variables
head(data_clean[, c(
  "AppointmentDay",
  "AppointmentDayOfWeek",
  "AppointmentMonth",
  "AppointmentWeek",
  "WaitingDays"
)])

```

# 15. Appointment Demand by Day of the Week

The number of appointments is examined across different days of the week to identify patterns in healthcare demand.

```{r demand-by-day}

# Count appointments by day of the week
day_demand <- table(data_clean$AppointmentDayOfWeek)

# Display the appointment counts
day_demand


```
# 16. Visualization of Appointment Demand by Day

The distribution of appointments across the days of the week is visualized to identify the days with higher healthcare demand and potential variations in patient demand throughout the week.

```{r demand-day-plot}

# Set the days in the correct weekly order
day_demand_df <- data.frame(
  Day = c("Monday", "Tuesday", "Wednesday",
          "Thursday", "Friday", "Saturday"),
  Appointments = c(
    day_demand["Monday"],
    day_demand["Tuesday"],
    day_demand["Wednesday"],
    day_demand["Thursday"],
    day_demand["Friday"],
    day_demand["Saturday"]
  )
)

# Create the bar chart
barplot(
  day_demand_df$Appointments,
  names.arg = day_demand_df$Day,
  main = "Appointment Demand by Day of the Week",
  xlab = "Day of the Week",
  ylab = "Number of Appointments",
  las = 2
)
```

**Interpretation:** The visualization shows that appointment demand varies across the days of the week. Wednesday records the highest number of appointments, closely followed by Tuesday. Monday also shows relatively high demand, while Friday and Thursday have fewer appointments. Saturday has an exceptionally low number of appointments. This suggests that patient demand is concentrated mainly during weekdays, particularly in the middle of the week. These findings may help the campus health centre plan staff availability and allocate resources more effectively during high-demand days.

# 17. Appointment Demand by Month

The number of appointments is examined across different months to identify seasonal patterns in healthcare demand.

```{r demand-by-month}

# Count appointments by month
month_demand <- table(data_clean$AppointmentMonth)

# Display the appointment counts
month_demand
```
# 18. Visualization of Appointment Demand by Month

The distribution of appointments across the available months is visualized to examine patterns in healthcare demand.

```{r demand-month-plot}

# Convert the monthly appointment counts into a data frame
month_demand_df <- as.data.frame(month_demand)

# Rename the columns
names(month_demand_df) <- c("Month", "Appointments")

# Set the months in chronological order
month_order <- c("April", "May", "June")

# Arrange the data according to the desired order
month_demand_df <- month_demand_df[
  match(month_order, month_demand_df$Month),
]

# Create the bar chart
barplot(
  month_demand_df$Appointments,
  names.arg = month_demand_df$Month,
  main = "Appointment Demand by Month",
  xlab = "Month",
  ylab = "Number of Appointments"
)
```
### Interpretation

The monthly analysis shows that the highest number of appointments occurred in May, followed by June, while April had the fewest appointments. However, these differences should be interpreted carefully because the dataset covers only a limited period and the number of observations available for each month is not equal. Therefore, the results indicate the distribution of appointments in this dataset rather than a definitive seasonal demand pattern.

# 19. Appointment Attendance and No-Show Analysis

The appointment attendance outcome is examined to understand the proportion of patients who attended their scheduled appointments and those who did not.

```{r no-show-analysis}

# Count attended and missed appointments
no_show_counts <- table(data_clean$No.show)

# Display the counts
no_show_counts

# Calculate the percentage of each outcome
no_show_percent <- prop.table(no_show_counts) * 100

# Display the percentages
round(no_show_percent, 2)
```
### Interpretation

The analysis shows that 79.81% of appointments were attended, while 20.19% resulted in a no-show. This means that approximately one out of every five scheduled appointments was missed. The relatively high proportion of no-shows indicates that missed appointments could be an important area for the campus health centre to address. Predictive analysis can therefore be used to identify factors associated with no-shows and support better appointment management.

# 20. SMS Reminder and No-Show Analysis

The relationship between receiving an SMS reminder and appointment attendance is examined to determine whether reminder messages are associated with no-show behaviour.

```{r sms-no-show}

# Create a cross-tabulation of SMS reminders and appointment outcomes
sms_no_show <- table(
  data_clean$SMS_received,
  data_clean$No.show
)

# Display the counts
sms_no_show

# Calculate percentages within each SMS group
round(prop.table(sms_no_show, margin = 1) * 100, 2)
```
### Interpretation

The analysis shows that 16.70% of appointments without an SMS reminder resulted in a no-show, compared with 27.57% among appointments where an SMS reminder was received. Although the no-show rate is higher among patients who received an SMS reminder, this result does not establish that SMS reminders cause patients to miss appointments. Other factors, such as waiting time, age, or appointment scheduling patterns, may influence both SMS receipt and attendance. Further analysis is therefore required before making a business decision about SMS reminders.

# 21. Age and No-Show Analysis

The relationship between patient age and appointment attendance is examined to determine whether no-show behaviour differs across age groups.

```{r age-no-show}

# Create age groups
data_clean$AgeGroup <- cut(
  data_clean$Age,
  breaks = c(-1, 17, 30, 45, 60, 100),
  labels = c("0-17", "18-30", "31-45", "46-60", "61-100")
)

# Calculate no-show percentage for each age group
age_no_show <- prop.table(
  table(data_clean$AgeGroup, data_clean$No.show),
  margin = 1
) * 100

# Display the percentages
round(age_no_show, 2)
```
**Interpretation:** The analysis shows that no-show behaviour varies across age groups. The 18–30 age group has the highest no-show rate at 24.61%, while the 61–100 age group has the lowest no-show rate at 15.20%. Overall, the no-show rate tends to decrease as age increases. This suggests that younger patients may require greater attention when planning appointment reminders and attendance management. However, age alone may not explain no-show behaviour, so other factors should also be considered in the predictive analysis.

# 22. Visualization of No-Show Rate by Age Group

The no-show rates across different age groups are visualized to better understand how appointment attendance varies with age.

```{r age-no-show-plot}

# Extract no-show percentages
age_no_show_percent <- age_no_show[, "Yes"]

# Create the bar chart
barplot(
  age_no_show_percent,
  names.arg = rownames(age_no_show),
  main = "No-Show Rate by Age Group",
  xlab = "Age Group",
  ylab = "No-Show Rate (%)",
  ylim = c(0, 30)
)
```
**Interpretation:** The visualization shows that the no-show rate is highest among the 18–30 age group, at 24.61%. The 61–100 age group has the lowest no-show rate, at 15.20%. Overall, the no-show rate generally decreases with increasing age, although the 0–17 and 31–45 groups have similar rates. This indicates that younger age groups, particularly 18–30, may require greater attention when planning appointment reminders and attendance management.

# 23. Waiting Time and No-Show Analysis

The relationship between the number of days between scheduling and the appointment date and patient no-show behaviour is examined to determine whether longer waiting periods are associated with missed appointments.

```{r waiting-no-show}

# Create waiting time groups
data_clean$WaitingGroup <- cut(
  data_clean$WaitingDays,
  breaks = c(-1, 2, 7, 14, 30, 179),
  labels = c("0-2 days", "3-7 days", "8-14 days", "15-30 days", "31+ days")
)

# Calculate no-show percentage for each waiting time group
waiting_no_show <- prop.table(
  table(data_clean$WaitingGroup, data_clean$No.show),
  margin = 1
) * 100

# Display the percentages
round(waiting_no_show, 2)
```
**Interpretation:** The analysis shows a clear relationship between waiting time and no-show behaviour. Appointments scheduled within 0–2 days have a relatively low no-show rate of 8.92%, while appointments with a waiting period of 31 days or more have a no-show rate of 33.00%. The no-show rate generally increases as the waiting period becomes longer. This suggests that longer waiting periods may be associated with a higher likelihood of missed appointments. Therefore, waiting time can be considered an important factor for appointment management and predictive analysis.

# 24. Visualization of No-Show Rate by Waiting Time

The no-show rates across different waiting time groups are visualized to better understand whether longer waiting periods are associated with missed appointments.

```{r waiting-no-show-plot}

# Extract no-show percentages
waiting_no_show_percent <- waiting_no_show[, "Yes"]

# Create the bar chart
barplot(
  waiting_no_show_percent,
  names.arg = rownames(waiting_no_show),
  main = "No-Show Rate by Waiting Time",
  xlab = "Waiting Time",
  ylab = "No-Show Rate (%)",
  ylim = c(0, 40)
)
```
**Interpretation:** The visualization clearly shows that the no-show rate increases as the waiting time becomes longer. Appointments with a waiting period of 0–2 days have the lowest no-show rate at 8.92%, while appointments with a waiting period of 31 days or more have the highest rate at 33.00%. This pattern suggests that longer waiting periods are associated with a greater likelihood of missed appointments. Therefore, reducing excessive waiting times or providing additional reminders for appointments scheduled far in advance may help improve attendance.

# 25. Gender and No-Show Analysis

The relationship between patient gender and appointment attendance is examined to determine whether no-show behaviour differs between gender groups.

```{r gender-no-show}

# Calculate attendance and no-show percentages by gender
gender_no_show <- prop.table(
  table(data_clean$Gender, data_clean$No.show),
  margin = 1
) * 100

# Display the percentages
round(gender_no_show, 2)
```
**Interpretation:** The analysis shows that the no-show rates for female and male patients are very similar. Female patients have a no-show rate of 20.31%, compared with 19.96% for male patients. The difference of only 0.35 percentage points suggests that gender has a relatively small association with no-show behaviour in this dataset. Therefore, gender alone may not be a strong factor for predicting missed appointments, although it can still be included as a potential predictor in the predictive model.

# 26. Visualization of No-Show Rate by Gender

The no-show rates for male and female patients are visualized to compare appointment attendance behaviour between the two gender groups.

```{r gender-no-show-plot}

# Extract no-show percentages
gender_no_show_percent <- gender_no_show[, "Yes"]

# Create the bar chart
barplot(
  gender_no_show_percent,
  names.arg = rownames(gender_no_show),
  main = "No-Show Rate by Gender",
  xlab = "Gender",
  ylab = "No-Show Rate (%)",
  ylim = c(0, 25)
)
```
**Interpretation:** The visualization shows that the no-show rates for female and male patients are almost identical. The no-show rate is 20.31% for female patients and 19.96% for male patients. The very small difference suggests that gender has little association with no-show behaviour in this dataset. Therefore, gender is unlikely to be a strong factor for predicting missed appointments on its own, although it may still be considered along with other variables in the predictive model.

# 27. Preparing Data for Predictive Modeling

The dataset is prepared for predictive modeling by converting the appointment outcome into a numerical format and selecting relevant variables that may help predict whether an appointment will result in a no-show.

```{r model-preparation}

# Convert no-show outcome into numerical form
data_clean$NoShow <- ifelse(data_clean$No.show == "Yes", 1, 0)

# Convert categorical variables to factors
data_clean$Gender <- as.factor(data_clean$Gender)
data_clean$AppointmentDayOfWeek <- as.factor(data_clean$AppointmentDayOfWeek)
data_clean$AppointmentMonth <- as.factor(data_clean$AppointmentMonth)

# Check the new target variable
table(data_clean$NoShow)

# Display the variables that will be considered for prediction
names(data_clean)
```

# 28. Splitting Data into Training and Testing Sets

The cleaned dataset is divided into training and testing sets. The training set is used to build the predictive model, while the testing set is used to evaluate how well the model performs on unseen data.

```{r train-test-split}

# Set a seed so that the same split can be reproduced
set.seed(123)

# Select relevant variables for the predictive model
model_data <- data_clean[, c(
  "NoShow",
  "Age",
  "Gender",
  "WaitingDays",
  "SMS_received",
  "Scholarship",
  "Hipertension",
  "Diabetes",
  "Alcoholism",
  "Handcap",
  "AppointmentDayOfWeek"
)]

# Remove records with missing values
model_data <- na.omit(model_data)

# Create random training and testing indices
train_index <- sample(
  1:nrow(model_data),
  size = 0.80 * nrow(model_data)
)

# Create training and testing datasets
train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]

# Check the number of records in each dataset
nrow(train_data)
nrow(test_data)
```
# 29. Building the Logistic Regression Model

A logistic regression model is developed to predict whether a patient will miss an appointment. The model uses patient and appointment-related characteristics from the training dataset.

```{r logistic-model}

# Build the logistic regression model
logistic_model <- glm(
  NoShow ~ Age + Gender + WaitingDays + SMS_received +
    Scholarship + Hipertension + Diabetes +
    Alcoholism + Handcap + AppointmentDayOfWeek,
  data = train_data,
  family = binomial
)

# Display the model summary
summary(logistic_model)
```
# 30. Making Predictions on the Test Data

The trained logistic regression model is used to predict the probability that each appointment in the testing dataset will result in a no-show.

```{r model-predictions}

# Predict the probability of a no-show
predicted_probability <- predict(
  logistic_model,
  newdata = test_data,
  type = "response"
)

# View the first few predicted probabilities
head(predicted_probability)
```
# 31. Converting Probabilities into Predictions

The predicted probabilities are converted into binary predictions using a threshold of 0.5. A probability of 0.5 or higher is classified as a predicted no-show, while a probability below 0.5 is classified as an expected attendance.

```{r predicted-class}

# Convert probabilities into Yes/No predictions
predicted_class <- ifelse(
  predicted_probability >= 0.5,
  1,
  0
)

# View the first few predictions
head(predicted_class)
```
# 32. Evaluating the Predictions Using a Confusion Matrix

A confusion matrix is used to compare the predicted appointment outcomes with the actual outcomes in the testing dataset. It shows how many appointments were correctly or incorrectly classified by the logistic regression model.

```{r confusion-matrix}

# Create the confusion matrix
confusion_matrix <- table(
  Actual = test_data$NoShow,
  Predicted = predicted_class
)

confusion_matrix
```
# 33. Calculating Model Accuracy

Model accuracy measures the proportion of test appointments that were classified correctly by the logistic regression model.

```{r model-accuracy}

# Calculate model accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)

accuracy
```
# 34. Calculating Sensitivity for No-Show Prediction

Sensitivity measures the ability of the model to correctly identify appointments that actually result in a no-show. This is an important measure because identifying potential no-shows is a key objective of the predictive analysis.

```{r model-sensitivity}

# Calculate sensitivity for no-show prediction
true_positives <- confusion_matrix["1", "1"]
false_negatives <- confusion_matrix["1", "0"]

sensitivity <- true_positives / (true_positives + false_negatives)

sensitivity
```
# 35. Calculating Specificity

Specificity measures the ability of the model to correctly identify appointments that are actually attended. It shows how well the model avoids incorrectly classifying attended appointments as no-shows.

```{r model-specificity}

# Calculate specificity
true_negatives <- confusion_matrix["0", "0"]
false_positives <- confusion_matrix["0", "1"]

specificity <- true_negatives / (true_negatives + false_positives)

specificity
```
# 36. Calculating Precision for No-Show Prediction

Precision measures how many of the appointments predicted as no-shows were actually no-shows. This helps evaluate how reliable the model's no-show predictions are.

```{r model-precision}

# Calculate precision
precision <- true_positives / (true_positives + false_positives)

precision
```
# 37. Calculating F1-Score

The F1-score combines precision and sensitivity into a single measure. It is particularly useful when the classes are imbalanced, as in this dataset where no-shows are less common than attended appointments.

```{r model-f1}

# Calculate F1-score
f1_score <- 2 * (precision * sensitivity) /
  (precision + sensitivity)

f1_score
```
# 38. ROC Curve and Area Under the Curve

The Receiver Operating Characteristic (ROC) curve is used to evaluate the ability of the logistic regression model to distinguish between attended and missed appointments. The Area Under the Curve (AUC) provides a summary measure of this discriminatory ability.

```{r roc-auc}

# Load the pROC package
library(pROC)

# Create the ROC curve
roc_curve <- roc(
  test_data$NoShow,
  predicted_probability
)

# Display the ROC curve
plot(
  roc_curve,
  main = "ROC Curve for Logistic Regression Model"
)

# Calculate the AUC
auc_value <- auc(roc_curve)

auc_value
```
**Interpretation:** The ROC curve lies above the diagonal reference line, indicating that the logistic regression model has some ability to distinguish between attended and missed appointments. The Area Under the Curve (AUC) is 0.6614, which indicates modest discriminatory performance. Therefore, the model can identify some differences between appointments that are likely to be attended and those likely to result in a no-show, but its predictive ability is limited. Further improvement in the model or adjustment of the classification threshold may be required to identify potential no-shows more effectively.

# 39. Finding an Improved Classification Threshold

The default classification threshold of 0.5 results in very low sensitivity for detecting no-shows. An alternative threshold is therefore examined using the ROC curve. The Youden method identifies a threshold that provides a balance between sensitivity and specificity.

```{r optimal-threshold}

# Find the threshold that gives the best balance
# between sensitivity and specificity
best_threshold <- coords(
  roc_curve,
  "best",
  ret = c("threshold", "sensitivity", "specificity"),
  best.method = "youden"
)

best_threshold
```
# 40. Predictions Using the Improved Threshold

The improved threshold identified from the ROC analysis is used to classify appointments as either expected attendance or potential no-show.

```{r improved-predictions}

# Extract the optimal threshold as a numeric value
optimal_threshold <- best_threshold$threshold

# Create new predictions using the improved threshold
improved_predicted_class <- ifelse(
  predicted_probability >= optimal_threshold,
  1,
  0
)

# View the first few predictions
head(improved_predicted_class)
```
# 41. Confusion Matrix Using the Improved Threshold

The predictions generated using the improved threshold are compared with the actual outcomes to evaluate whether the model is better at identifying no-show appointments.

```{r improved-confusion-matrix}

# Create the new confusion matrix
improved_confusion_matrix <- table(
  Actual = test_data$NoShow,
  Predicted = improved_predicted_class
)

improved_confusion_matrix
```
# 42. Sensitivity Using the Improved Threshold

Sensitivity measures the proportion of actual no-show appointments that are correctly identified by the model. This is particularly important because the main objective is to identify appointments that are likely to be missed.

```{r improved-sensitivity}

# Extract values from the improved confusion matrix
true_positives_improved <- improved_confusion_matrix["1", "1"]
false_negatives_improved <- improved_confusion_matrix["1", "0"]

# Calculate sensitivity
improved_sensitivity <- true_positives_improved /
  (true_positives_improved + false_negatives_improved)

improved_sensitivity
```
# 43. Specificity Using the Improved Threshold

Specificity measures the proportion of appointments that were actually attended and were correctly identified by the model as expected attendance.

```{r improved-specificity}

# Extract true negatives and false positives
true_negatives_improved <- improved_confusion_matrix["0", "0"]
false_positives_improved <- improved_confusion_matrix["0", "1"]

# Calculate specificity
improved_specificity <- true_negatives_improved /
  (true_negatives_improved + false_positives_improved)

improved_specificity
```
# 44. Precision Using the Improved Threshold

Precision measures the proportion of appointments predicted as no-shows that were actually no-shows. It helps determine how reliable the model's no-show predictions are.

```{r improved-precision}

# Extract true positives and false positives
true_positives_improved <- improved_confusion_matrix["1", "1"]
false_positives_improved <- improved_confusion_matrix["0", "1"]

# Calculate precision
improved_precision <- true_positives_improved /
  (true_positives_improved + false_positives_improved)

improved_precision
```
# 45. F1-Score Using the Improved Threshold

The F1-score combines precision and sensitivity into a single measure. It is useful for evaluating the model when the classes are imbalanced and both correctly identifying no-shows and avoiding incorrect predictions are important.

```{r improved-f1-score}

# Calculate the F1-score
improved_f1_score <- 2 * (
  (improved_precision * improved_sensitivity) /
  (improved_precision + improved_sensitivity)
)

improved_f1_score
```
# 46. Accuracy Using the Improved Threshold

The accuracy of the model is recalculated using the improved classification threshold to determine the overall proportion of correctly classified appointments.

```{r improved-accuracy}

# Calculate accuracy using the improved predictions
improved_accuracy <- sum(
  diag(improved_confusion_matrix)
) / sum(improved_confusion_matrix)

improved_accuracy
```
# 47. Comparison of Classification Thresholds

The performance of the logistic regression model is compared using the default threshold of 0.5 and the improved threshold identified using the Youden method.

```{r threshold-comparison}

# Create a comparison table
threshold_comparison <- data.frame(
  Metric = c(
    "Accuracy",
    "Sensitivity",
    "Specificity",
    "Precision",
    "F1-Score"
  ),
  Default_Threshold = c(
    accuracy,
    sensitivity,
    specificity,
    precision,
    f1_score
  ),
  Improved_Threshold = c(
    improved_accuracy,
    improved_sensitivity,
    improved_specificity,
    improved_precision,
    improved_f1_score
  )
)

# Convert values to percentages
threshold_comparison$Default_Threshold <- round(
  threshold_comparison$Default_Threshold * 100, 2
)

threshold_comparison$Improved_Threshold <- round(
  threshold_comparison$Improved_Threshold * 100, 2
)

threshold_comparison
```

**Interpretation:** The comparison shows that changing the classification threshold from 0.5 to approximately 0.1908 substantially improves the model's ability to identify no-show appointments. Sensitivity increases from 1.69% to 62.36%, while the F1-score increases from 3.23% to 41.60%. However, accuracy and specificity decrease because more appointments are classified as potential no-shows. Since identifying potential no-shows is an important objective of the analysis, the improved threshold provides a more useful balance than the default threshold.

# 48. Business Interpretation of the Predictive Model

The predictive analysis was conducted to identify appointments that may be at higher risk of resulting in a no-show. The logistic regression model achieved an AUC of 0.6614, indicating that the model has a modest ability to distinguish between appointments that are attended and those that are missed.

Using the default classification threshold of 0.5, the model achieved an accuracy of 79.36%, but its sensitivity was only 1.69%. This means that the model correctly identified very few of the appointments that actually resulted in no-shows.

An improved classification threshold of approximately 0.1908 was therefore examined. At this threshold, sensitivity increased to 62.36%, meaning that the model was able to identify a substantially larger proportion of actual no-show appointments. The specificity was 64.91%, while the F1-score increased from 3.23% to 41.60%.

These results show that accuracy alone is not sufficient for evaluating a no-show prediction system because the dataset contains considerably more attended appointments than no-show appointments. For appointment management, identifying potential no-shows is an important objective, so sensitivity and F1-score provide useful additional measures of model performance.

The predictive model can therefore be used as a decision-support tool to flag appointments that have a higher predicted probability of being missed. The health centre could use these predictions to prioritize reminder efforts or review appointment scheduling practices. However, the predictions should support staff decision-making rather than automatically determine whether an appointment should be cancelled or treated differently.

# 49. Key Findings from the Analysis

The analysis produced several important findings related to appointment demand and patient attendance.

1. **Appointment demand varies across the week:**  
   Wednesday recorded the highest number of appointments, followed closely by Tuesday, while Saturday had very few appointments. This indicates that appointment demand is concentrated mainly on weekdays.

2. **Appointment demand varies across the available months:**  
   May recorded the highest number of appointments in the dataset, followed by June and April. However, these results should be interpreted cautiously because the dataset covers a limited period and the number of observations available for each month is not equal.

3. **No-shows represent a significant proportion of appointments:**  
   Approximately 20.19% of appointments resulted in a no-show, meaning that around one in every five scheduled appointments was missed.

4. **Waiting time is associated with no-show behaviour:**  
   The no-show rate increased from 8.92% for appointments with a waiting period of 0–2 days to 33.00% for appointments with a waiting period of 31 days or more. This was one of the clearest patterns observed in the analysis.

5. **No-show rates vary across age groups:**  
   The 18–30 age group had the highest observed no-show rate at 24.61%, while the 61–100 age group had the lowest at 15.20%.

6. **Gender showed little difference in no-show rates:**  
   Female patients had a no-show rate of 20.31%, compared with 19.96% for male patients. The small difference suggests that gender alone has limited association with no-show behaviour in this dataset.

7. **The predictive model has limited but useful discriminatory ability:**  
   The logistic regression model achieved an AUC of 0.6614. This indicates that the model can identify some differences between appointments likely to be attended and those likely to result in a no-show, although its predictive ability is limited.

8. **Classification threshold has a major effect on model performance:**  
   At the default threshold of 0.5, sensitivity was only 1.69%. After using an improved threshold of approximately 0.1908, sensitivity increased to 62.36% and the F1-score increased from 3.23% to 41.60%. This demonstrates the importance of selecting a classification threshold according to the practical objective of the prediction system.
   

# 50. Business Recommendations

Based on the demand analysis, no-show analysis, and predictive modelling results, the following recommendations can be considered for improving appointment management at a campus health centre.

### 1. Plan staff availability according to appointment demand

Appointment demand was higher on Tuesday and Wednesday compared with the other days of the week. The health centre can use historical appointment patterns to plan staff availability and supporting resources during periods of relatively higher demand.

### 2. Pay attention to appointments with longer waiting periods

The analysis showed that no-show rates increased as the waiting period became longer. The health centre can therefore monitor appointments scheduled far in advance and consider additional communication or reminder strategies for such appointments.

### 3. Use predictive risk flags for appointment management

The logistic regression model can be used to calculate a predicted probability of a no-show for each appointment. Appointments above the selected threshold can be flagged as potential high-risk appointments.

These flags can help health-centre staff prioritize reminder efforts rather than treating every appointment in exactly the same way.

### 4. Use targeted reminders rather than relying only on general reminders

The analysis showed differences in no-show behaviour across waiting-time and age groups. These patterns can be used to design more targeted appointment-management strategies.

For example, appointments with longer waiting periods could receive additional reminders closer to the appointment date.

### 5. Use the model as a decision-support tool

The predictive model should be used to support staff decisions rather than automatically cancelling or rejecting appointments. Staff can consider the model's prediction together with practical information such as appointment type, urgency, and patient circumstances.

### 6. Continuously monitor appointment data

The health centre should regularly analyse new appointment data to identify changes in demand, attendance, and no-show patterns. As more campus-specific data becomes available, the predictive model can also be retrained and evaluated using the health centre's own historical records.

### 7. Collect campus-specific data for future improvement

The current analysis uses a publicly available healthcare appointment dataset adapted as a proxy for the campus health-centre scenario. A real implementation should collect campus-specific information such as appointment type, student/staff category, reason for visit, booking time, waiting time, previous attendance behaviour, and appointment cancellation status. This would allow a future model to provide more relevant predictions for the actual campus health centre.

# 51. Proposed Real-Life Analytics Solution

The analysis can be developed into a practical appointment-management system for a campus health centre. The proposed solution would use historical appointment data to understand demand patterns and identify appointments that may have a higher probability of resulting in a no-show.

### Step 1: Collect appointment data

The health centre can maintain a digital record of appointments containing information such as:

- Appointment date and time
- Booking date
- Waiting time between booking and appointment
- Student or staff category
- Age group
- Appointment type
- Previous attendance history
- Reminder status
- Cancellation status
- Whether the appointment was attended or missed

### Step 2: Analyse appointment demand

Historical appointment data can be analysed to identify:

- Days with higher appointment demand
- Periods with lower or higher demand
- Changes in appointment volume over time
- Patterns in cancellations and no-shows

This information can help the health centre plan staff availability and resources according to expected demand.

### Step 3: Calculate no-show risk

The predictive model can use appointment and patient-related information to calculate the probability that an appointment will result in a no-show.

For example, the model could produce an output such as:

| Appointment | Predicted No-Show Probability | Risk Flag |
|---|---:|---|
| A001 | 8% | Lower risk |
| A002 | 14% | Lower risk |
| A003 | 27% | Potential risk |
| A004 | 41% | Potential risk |

The exact risk categories would need to be determined using future campus-specific data and operational requirements.

### Step 4: Prioritize reminder efforts

Appointments identified as having a higher predicted probability of no-show can be given additional attention. For example, the health centre could send an additional reminder or confirmation request before the appointment.

This allows staff to focus limited resources on appointments that may require additional follow-up rather than applying the same intervention to every appointment.

### Step 5: Monitor the results

After implementing the system, the health centre can compare predicted outcomes with actual attendance.

The model can then be evaluated regularly using measures such as:

- Sensitivity
- Specificity
- Precision
- F1-score
- AUC

This would allow the health centre to determine whether the prediction system continues to provide useful information.

### Proposed workflow

The overall proposed system can therefore be represented as:

**Appointment Data → Data Cleaning → Demand Analysis → No-Show Prediction → Risk Flagging → Targeted Reminder → Attendance Monitoring → Model Improvement**

The proposed system would function as a decision-support system for the campus health centre. It would not automatically cancel appointments or make decisions about patients. Instead, it would provide staff with analytical information that can support appointment planning and reminder management.

The current project demonstrates this workflow using a publicly available healthcare appointment dataset adapted as a proxy for the campus health-centre scenario. A future implementation should retrain and validate the model using actual campus health-centre data before it is used operationally.

# 52. Limitations of the Study

Although the analysis provides useful insights into appointment demand and no-show behaviour, several limitations should be considered when interpreting the results.

### 1. Use of a proxy dataset

The dataset used in this project is a publicly available medical appointment dataset and was adapted to demonstrate a campus health-centre scenario. It does not represent actual appointment data from the campus health centre. Therefore, the findings should be considered a demonstration of the proposed analytics approach rather than direct evidence about the behaviour of students or staff at the campus.

### 2. Limited time coverage

The dataset covers a relatively short period of time. In addition, the number of appointments available for each month is not equal. Therefore, differences in monthly appointment volume cannot be interpreted as a definitive seasonal pattern.

### 3. Limited predictive performance

The logistic regression model achieved an AUC of 0.6614, indicating modest discriminatory ability. This means that the model can identify some differences between attended and missed appointments, but it cannot predict no-shows perfectly.

### 4. Trade-off caused by the classification threshold

The improved threshold increased sensitivity from 1.69% to 62.36%, allowing the model to identify a larger proportion of actual no-shows. However, specificity and overall accuracy decreased. This means that some appointments predicted as potential no-shows would actually be attended.

### 5. Missing campus-specific variables

The dataset does not contain several variables that could be important in a campus health-centre environment. Examples include student or staff status, appointment type, reason for visit, academic schedule, examination periods, previous attendance history, and campus-specific appointment policies.

### 6. Predictive analysis does not establish causation

The relationships observed between variables and no-show behaviour should not be interpreted as proof that one variable directly causes another. For example, the higher no-show rate observed among appointments where an SMS reminder was received does not mean that receiving an SMS causes patients to miss appointments. Other factors may influence both variables.

### 7. Future validation is required

Before the proposed system is used in a real campus health centre, the model should be trained and tested using actual campus-specific appointment data. Its performance should also be monitored over time and the model should be updated when sufficient new data becomes available.

Despite these limitations, the project demonstrates how business analytics and predictive modelling can be applied to appointment demand and no-show management and provides a framework that can be adapted to a real campus health-centre environment.

# 53. Conclusion

This project analysed appointment demand and no-show behaviour using a publicly available healthcare appointment dataset adapted to a campus health-centre scenario. The analysis followed a business analytics approach involving data cleaning, exploratory analysis, feature engineering, and predictive modelling.

The demand analysis showed that appointment volume varied across different days of the week, with higher appointment volumes observed during the middle of the week. The no-show analysis showed that approximately 20.19% of appointments were missed. Waiting time showed a clear association with no-show behaviour, with longer waiting periods generally corresponding to higher no-show rates.

A logistic regression model was developed to predict whether an appointment would result in a no-show. The model achieved an AUC of 0.6614, indicating modest ability to distinguish between attended and missed appointments. The default classification threshold of 0.5 resulted in very low sensitivity of 1.69%. By using an improved threshold of approximately 0.1908, sensitivity increased to 62.36% and the F1-score increased from 3.23% to 41.60%. This demonstrates that classification threshold selection is important when the objective is to identify potential no-show appointments.

Based on these findings, a real-life appointment-management solution was proposed in which historical appointment data could be used to analyse demand, predict potential no-shows, prioritize reminder efforts, and continuously monitor attendance patterns.

The current project demonstrates the potential of predictive analytics as a decision-support approach for healthcare appointment management. However, the model should be retrained and validated using actual campus health-centre data before being considered for real-world implementation.

Overall, the project demonstrates how Business Analytics can transform appointment data into actionable information that can support demand planning, appointment management, and more efficient use of healthcare resources.

